Mixture Models for Single-Cell Assays with Application to Vaccine Studies: Supplementary Materials

### 2013-05-14 SupplementaryMaterialMIMOSA.tex
Updated notation from _{s,u} to ^{(s,u)}.

We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$. In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1). Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)$ but requires computing \[
Z(\alpha_u, \beta_u, \alpha_s, \beta_s)=\int_{0}^1{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}\int_{p_u}^1 p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}dp_sdp_u.

@@

\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.

@@

In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used. Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation. A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration. Indeed, we can write

@@

\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)=\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\int_{0}^1\frac{{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}}{\mathrm{B}(\alpha_u,\beta_u)}[1-F_{\alpha_s,\beta_s}(p_u)]dp_u

@@

where $F_{\alpha_s,\beta_s}$ is the cumulative distribution function of a beta random variable with parameters $\alpha_s$ and $\beta_s$. Using this identity, it can be seen that $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ can be approximated by

@@

\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)\approx\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\sum_{k=1}^K[1-F_{\alpha_s,\beta_s}(X_k)]

@@

where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha_s$, $\beta_s$ and $K$ is the number of terms used in the Monte Carlo approximation. This approximation works relatively well with our EM implementation and does not significantly increase the computing time.

@@

can be obtained when $\alpha_s$ and $\beta_s$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,

@@

F_{\alpha_s,\beta_s}(p_u)=\sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^jp_u^{\beta_s+\alpha_s-j}.

@@

Z(\alpha_u, \beta_u, \alpha_s, \beta_s)&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\int_{0}^1(1-p_u)^{\beta_u-1+j}p_u^{\alpha_u-1+\beta_s+\alpha_s-j}dp_u\\
&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\mathrm{B}(\beta_u+j)\mathrm{B}(\alpha_u+\beta_s+\alpha_s-j).

@@

Typically, in ICS data $\alpha_s$ is relatively small leading to relatively few terms in the sum. However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha_s$ and $\beta_s$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha_s$ is less than one) and can render mixing in the MCMC more difficult. In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model. In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is greater than that of the stimulation sample, $p_s$. Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$. In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.

\subsection{Computational details for the beta-binomial model}
\noindent\textbf{Marginal likelihood derivations}\\
For a given subject $i$, the null marginal likelihood is obtained after integrating out the prior from the likelihood for model ${\cal M}_0$, as follows,
\[
\begin{split}
\mathrm{L_0}(\alpha_u,\beta_u|n_{si},n_{ui}) =&\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u)\pi(p_u|\alpha_u,\beta_u)\mathrm{d}p_u\\
=& \int_0^1Pr(n_s|p_u)Pr(n_u|p_u)\pi(p_u|\alpha_u,\beta_u) \mathrm{d}p_u\\
=&\int_0^1\binom{N_s}{n_s}p_u^{n_s}(1-p_u)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \mathrm{d}p_u\\
=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\int_0^1 p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u-n_s-n_u+\beta_u-1} \mathrm{d}p_u\\
=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u-n_s-n_u+\beta_0)}{\mathrm{B}(\alpha_u,\beta_u)}.
\end{split}
\]

\noindent Similarly, the alternative marginal likelihood for a given subject is defined as,
\[
\begin{split}
\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{si},n_{ui}) =&\int_0^1\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u,p_s)\pi(p_u,p_s|\alpha_u,\beta_u,\alpha_s,\beta_s)\mathrm{d}p_s\mathrm{d}p_u \\
=&\int_0^1\int_0^1\binom{N_s}{n_s}p_s^{n_s}(1-p_s)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \cdot\\
&\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1} \mathrm{d}p_s\mathrm{d}p_u\\
=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\int_0^1 p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u-n_u+\beta_u-1} \cdot\\
&\int_0^1 p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s-n_s+\beta_s-1}\mathrm{d}p_s\mathrm{d}p_u \\
=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_u+\alpha_u,N_u-n_u+\beta_u)\mathrm{B}(n_s+\alpha_s,N_s-n_s+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}.
\end{split}
\]
%and taking the product over subjects leads to
%\[
%\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_{s},\mathbf{n}_{u}) = \prod_{i=1}^P \binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}
%\]

\noindent\textbf{MCMC algorithm}\\
In what follows, we use $(x|y)$ to denote the conditional distribution of $x$ given $y$. In particular, we use $(x|\cdots)$ to denote the distribution of $x$ conditional on everything else in the model. Our MCMC algorithms cycles through the following steps:
\begin{enumerate}
\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc}.
\item Update $w$ by Gibbs sampling using the full conditional,
\[
(w|\cdots)\sim \mathrm{Beta}(\sum_i z_i,\sum_i (1-z_i)).
\]
\item for each $i$, update $z_i$ by Gibbs sampling using the following full conditional,
\[
(z_i|\cdots)\sim\mathrm{B}(1,p_i).
\]
where,
\[
p_i=\frac{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s| n_{ui},n_{si})}{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{ui},n_{si})+(1-w)\cdot\mathrm{L_0}(\alpha_u,\beta_u|n_{ui},n_{si})}.
\]
\end{enumerate}
For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, (e.g. $\alpha_u$, below),
\[
\begin{split}
\frac{\mathrm{L}(\alpha^{\text{new}}_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha^{\text{new}}_u)}{\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha_u)}\\
\end{split}
\]
where $L$ is the complete marginal likelihood conditional on $\mathbf{z}$ defined as,
\[
\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s,\mathbf{z})=\prod_i\mathrm{L}_{z_i}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s)
\]
and $\pi$ is the prior distribution of $\alpha_u$. The obvious changes in the above expression are made for the acceptance ratios of $\alpha_s,\beta_s,\beta_u$. In our case each parameter has the same exponential prior with mean $1,000$

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.
For a given subject, the null marginal likelihood is defined as
\[
\begin{split}
\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui})=&\idotsint\mathrm{Pr}(\mathbf{n}_{si},\mathbf{n}_{ui}|\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\idotsint \frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\prod_k p^{n_{sik}}_{uk} \frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\prod_k p^{n_{uik}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sik}+n_{uik}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
\end{split}
\]
%The product over all $P$ observations gives the marginal likelihood,
%\[
%\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
%\]

>>>>

We can define a model where we constrain the stimulated proportions under the alternative model such that $p^{(s)}>p^{(u)}$. In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1). Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})$ but requires computing \[
Z(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})=\int_{0}^1{p^{(u)}}^{\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}\int_{p^{(u)}}^1 p^{(s)\alpha^{(s)}-1}(1-p^{(s)})^{\beta^{(s)}-1}dp^{(s)}dp^{(u)}.

@@

\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\frac{Z(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)},n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{Z(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)})}.

@@

In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used. Let us denote by $\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})$ the approximation. A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration. Indeed, we can write

@@

\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})=\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})\int_{0}^1\frac{{p^{(u)}}^{\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}[1-F_{\alpha^{(s)},\beta^{(s)}}(p^{(u)})]dp^{(u)}

@@

where $F_{\alpha^{(s)},\beta^{(s)}}$ is the cumulative distribution function of a beta random variable with parameters $\alpha^{(s)}$ and $\beta^{(s)}$. Using this identity, it can be seen that $\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})$ can be approximated by

@@

\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})\approx\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{k=1}^K[1-F_{\alpha^{(s)},\beta^{(s)}}(X_k)]

@@

where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha^{(s)}$, $\beta^{(s)}$ and $K$ is the number of terms used in the Monte Carlo approximation. This approximation works relatively well with our EM implementation and does not significantly increase the computing time.

@@

can be obtained when $\alpha^{(s)}$ and $\beta^{(s)}$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,

@@

F_{\alpha^{(s)},\beta^{(s)}}(p^{(u)})=\sum_{j=\beta^{(s)}}^{\beta^{(s)}+\alpha^{(s)}-1} \frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}(1-p^{(u)})^jp^{(u)\beta^{(s)}+\alpha^{(s)}-j}.

@@

Z(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})&=\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{j=\beta^{(s)}}^{\beta^{(s)}+\alpha^{(s)}-1}\frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}\int_{0}^1(1-p^{(u)})^{\beta^{(u)}-1+j}p^{(u)\alpha^{(u)}-1+\beta^{(s)}+\alpha^{(s)}-j}dp^{(u)}\\
&=\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{j=\beta}^{(s)\beta^{(s)}+\alpha^{(s)}-1}\frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}\mathrm{B}(\beta^{(u)}+j)\mathrm{B}(\alpha^{(u)}+\beta^{(s)}+\alpha^{(s)}-j).

@@

Typically, in ICS data $\alpha^{(u)}$ is relatively small leading to relatively few terms in the sum. However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha^{(s)}$ and $\beta^{(s)}$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha^{(s)}$ is less than one) and can render mixing in the MCMC more difficult. In addition, even though the computation is exact and much faster for small values of $\alpha^{(s)}$, which is typically the case with ICS data, it is still more demanding than the unconstrained model. In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p^{(u)}$, is greater than that of the stimulation sample, $p^{(s)}$. Indeed, in the one-sided case, if $p^{(u)}>p^{(s)}$ the associated individual should be a non-responder and thus $z_i=0$. In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.

\subsection{Computational details for the beta-binomial model}
\noindent\textbf{Marginal likelihood derivations}\\
For a given subject $i$, the null marginal likelihood is obtained after integrating out the prior from the likelihood for model ${\cal M}_0$, as follows,
\[
\begin{split}
\mathrm{L_0}(\alpha^{(u)},\beta^{(u)}|n^{(s)}_{i},n^{(u)}_{i}) =&\int_0^1 \mathrm{Pr}(n^{(s)}_{i},n^{(u)}_{i}|p^{(u)})\pi(p^{(u)}|\alpha^{(u)},\beta^{(u)})\mathrm{d}p^{(u)}\\
=& \int_0^1Pr(n^{(s)}|p^{(u)})Pr(n^{(u)}|p^{(u)})\pi(p^{(u)}|\alpha^{(u)},\beta^{(u)}) \mathrm{d}p^{(u)}\\
=&\int_0^1\binom{N^{(s)}}{n^{(s)}}p^{(s)n^{(s)}}(1-p^{(u)})^{N^{(s)}-n^{(s)}}\binom{N^{(u)}}{n^{(u)}}p^{(u)n^{(u)}}(1-p^{(u)})^{N^{(u)}-n^{(u)}}\cdot \\&\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}p^{(u)\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1} \mathrm{d}p^{(u)}\\
=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\int_0^1 p^{(u)n^{(s)}+n^{(u)}+\alpha^{(u)}-1}(1-p^{(u)})^{N^{(s)}+N^{(u)}-n^{(s)}-n^{(u)}+\beta^{(u)}-1} \mathrm{d}p^{(u)}\\
=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{\mathrm{B}(n^{(s)}+n^{(u)}+\alpha^{(u)},N^{(s)}+N^{(u)}-n^{(s)}-n^{(u)}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}.
\end{split}
\]

\noindent Similarly, the alternative marginal likelihood for a given subject is defined as,
\[
\begin{split}
\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|n^{(s)}_{i},n^{(u)}_{i}) =&\int_0^1\int_0^1 \mathrm{Pr}(n^{(s)}_{i},n^{(u)}_{i}|p^{(u)},p^{(s)})\pi(p^{(u)},p^{(s)}|\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)})\mathrm{d}p^{(s)}\mathrm{d}p^{(u)} \\
=&\int_0^1\int_0^1\binom{N^{(s)}}{n^{(s)}}p^{(s)n^{(s)}}(1-p^{(s)})^{N^{(s)}-n^{(s)}}\binom{N^{(u)}}{n^{(u)}}p^{(u)n^{(u)}}(1-p^{(u)})^{N^{(u)}-n^{(u)}} \cdot\\  &\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}p^{(u)\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}
\frac{1}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}p^{(s)\alpha^{(s)}-1}(1-p^{(s)})^{\beta^{(s)}-1} \mathrm{d}p^{(s)}\mathrm{d}p^{(u)}\\
=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\frac{1}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}\int_0^1 p^{(u)n^{(u)}+\alpha^{(u)}-1}(1-p^{(u)})^{N^{(u)}-n^{(u)}+\beta^{(u)}-1} \cdot\\
&\int_0^1 p^{(s)n^{(s)}+\alpha^{(s)}-1}(1-p^{(s)})^{N^{(s)}-n^{(s)}+\beta^{(s)}-1}\mathrm{d}p^{(s)}\mathrm{d}p^{(u)} \\
=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{\mathrm{B}(n^{(u)}+\alpha^{(u)},N^{(u)}-n^{(u)}+\beta^{(u)})\mathrm{B}(n^{(s)}+\alpha^{(s)},N^{(s)}-n^{(s)}+\beta^{(s)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})}.
\end{split}
\]
%and taking the product over subjects leads to
%\[
%\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_{s},\mathbf{n}_{u}) = \prod_{i=1}^P \binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}
%\]

\noindent\textbf{MCMC algorithm}\\
In what follows, we use $(x|y)$ to denote the conditional distribution of $x$ given $y$. In particular, we use $(x|\cdots)$ to denote the distribution of $x$ conditional on everything else in the model. Our MCMC algorithms cycles through the following steps:
\begin{enumerate}
\item Update each $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}$ and $\beta^{(s)}$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc}.
\item Update $w$ by Gibbs sampling using the full conditional,
\[
(w|\cdots)\sim \mathrm{Beta}(\sum_i z_i,\sum_i (1-z_i)).
\]
\item for each $i$, update $z_i$ by Gibbs sampling using the following full conditional,
\[
(z_i|\cdots)\sim\mathrm{B}(1,p_i).
\]
where,
\[
p_i=\frac{w\cdot\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}| n^{(u)}_{i},n^{(s)}_{i})}{w\cdot\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|n^{(u)}_{i},n^{(s)}_{i})+(1-w)\cdot\mathrm{L_0}(\alpha^{(u)},\beta^{(u)}|n^{(u)}_{i},n^{(s)}_{i})}.
\]
\end{enumerate}
For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, (e.g. $\alpha^{(u)}$, below),
\[
\begin{split}
\frac{\mathrm{L}(\alpha^{(u)\text{new}},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\cdots)\pi(\alpha^{(u)\text{new}})}{\mathrm{L}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\cdots)\pi(\alpha^{(u)})}\\
\end{split}
\]
where $L$ is the complete marginal likelihood conditional on $\mathbf{z}$ defined as,
\[
\mathrm{L}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{n}^{(u)},\mathbf{n}^{(s)},\mathbf{z})=\prod_i\mathrm{L}_{z_i}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{n}^{(u)},\mathbf{n}^{(s)})
\]
and $\pi$ is the prior distribution of $\alpha^{(u)}$. The obvious changes in the above expression are made for the acceptance ratios of $\alpha^{(s)},\beta^{(s)},\beta^{(u)}$. In our case each parameter has the same exponential prior with mean $1,000$

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.
For a given subject, the null marginal likelihood is defined as
\[
\begin{split}
\mathrm{L_0}(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})=&\idotsint\mathrm{Pr}(\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)})\pi(\mathbf{p}^{(u)}|\boldsymbol{\alpha}^{(u)})\mathrm{d}\mathbf{p}^{(u)}\\
=&\idotsint \frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\prod_k p^{(u)n^{(s)}_{ik}}_{k} \frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\prod_k p^{(u)n^{(u)}_{ik}}_{k}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}\prod_k p^{(u)\alpha^{(u)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\\
=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}\int_0^1 \prod_k p^{(u)n^{(s)}_{ik}+n^{(u)}_{ik}+\alpha^{(u)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\\
=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{\mathrm{B}(\mathbf{n}^{(s)}_{i}+\mathbf{n}^{(u)}+\boldsymbol{\alpha}^{(u)})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}
\end{split}
\]
%The product over all $P$ observations gives the marginal likelihood,
%\[
%\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
%\]

\noindent Similarly, the alternative marginal likelihood for a given subject is:
\[
\begin{split}
\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(u)}_{i},\mathbf{n}^{(s)}_{i})
%=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})\mathrm{B}(\boldsymbol{\alpha}^{(s)})}\idotsint\prod_k p^{(u)n^{(u)}_{ik}+\alpha^{(u)}_{k}-1}_{k}\idotsint \prod_k p^{(s)n^{(s)}_{ik}+\alpha^{(s)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\mathrm{d}\mathbf{p}^{(s)}\\
=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{\mathrm{B}(\mathbf{n}^{(u)}_{i}+\boldsymbol{\alpha}^{(u)})\mathrm{B}(\mathbf{n}^{(s)}_{i}+\boldsymbol{\alpha}^{(s)})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})\mathrm{B}(\boldsymbol{\alpha}^{(s)})}.
\end{split}
\]
\noindent\textbf{MCMC algorithm}\\
The MCMC algorithm for the Dirichlet-multinomial model is analogous to the beta-binomial, above. The parameter vectors $\boldsymbol{\alpha}^{(s)},\boldsymbol{\alpha}^{(u)}$ are updated component-wise:
\begin{enumerate}
\item  Update each $\alpha^{(s)}_{k}$, $\alpha^{(u)}_{k}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}.
\item Update $w$ by Gibbs sampling using the full conditional,
\[
(w|\cdots)\sim\mathrm{Beta}(\sum_i z_i,\sum_i(1-z_i))
\]
\item For each $i$, update $z_i$ using the full conditional,
\[
(z_i|\cdots)\sim \mathrm{B}(1,p_i)
\]
where
\[
p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})}
\]
For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha^{(u)}_{k}$, below):
\[
\frac{\mathrm{L}(\alpha^{(u)\text{new}}_{k},\boldsymbol{\alpha}^{(u)}_{\{-k\}},\boldsymbol{\alpha}^{(s)}|\cdots)\pi(\alpha^{(u)\text{new}}_{k})}{\mathrm{L}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\cdots)\pi(\alpha^{(u)}_{k})}\\
\]
